---
title: "Transit Oriented Development"
author: "Samriddhi Khare"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    theme: yeti
  
---
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    out.width = '100%',
    fig.retina =3
  )
```

## R Setup and Installing packages

loading packages and cansus key

```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)
library(tidyr)
library(ggplot2)
library(viridis)
library(prettydoc)

options(scipen=999)
options(tigris_class = "sf")

#these are functions developed for this class, so it takes you especially to the book to get those

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#fde725","#5ec962","#21918c","#3b528b","#440154")

census_api_key('bf2d507651b5a621dbadd44533fb4f3deaab26bf', overwrite = TRUE)

dir.create("output")

```


this is the dataset to use

```{r load_variables, cache=TRUE}

acs_variable_list.2020 <- load_variables(2020, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)

acs_variable_list.2016 <- load_variables(2016, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)
```

B01001B_001E black
B08101_025E people taking public transit to work
B24011_002E people working in business, science, and arts occupations
B25026_001E people living in households
B19013_001 median income
B25058_001E rent
B06012_002E under poverty line
B25097_001 mortgage

### Use `get_acs()` to get 2016 ACS data

Notice this returns "long" data - let's examine it

```{r data collection, include=TRUE, message=FALSE, warning=FALSE, results='hide'}
tracts16 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E",#pop
                        "B19013_001E", #med inc
                        "B25058_001E", #rent
                        "B24011_002E" ), #ppl in business science arts 
          year=2016, 
          state = 25,
          county=025, #using the boston/suffolk county FIPS code
          geometry=TRUE) %>% 
  st_transform('ESRI:102728')

```

```{r}
# making it wide from long

tracts16 <- 
  tracts16 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>% #spread makes it long to wide
  rename(TotalPop = B25026_001, 
         MedInc = B19013_001,
         MedRent = B25058_001, 
         BusArtsScience = B24011_002)

#creating new variable

tracts16 <- 
  tracts16 %>%
  mutate(IncSpentOnRentPCt = ifelse(MedInc > 0, MedRent / MedInc, 0),
         year = "2016") 
```

```{r results='hide'}

#doing the same thing for 2020

tracts20 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E",#pop
                        "B19013_001E", #med inc
                        "B25058_001E", #rent
                        "B24011_002E" ), #ppl in business science arts 
          year=2020, 
          state = 25,
          county=025, #using the boston/suffolk county FIPS code
          geometry=TRUE, output="wide")%>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         MedInc = B19013_001E,
         MedRent = B25058_001E, 
         BusArtsScience = B24011_002E)%>%
  dplyr::select(-NAME, -ends_with("M")) %>%
  mutate(IncSpentOnRentPCt = ifelse(MedInc > 0, MedRent / MedInc, 0),
         year = "2020")  
```

```{r}
#binding the data
allTracts <- rbind(tracts16,tracts20)
```

```{r name, results = 'hide'}

github_url <-"https://raw.githubusercontent.com/samriddhikhar3/Assignment2/main/data/MBTA_Systemwide_GTFS_Map.geojson"

# try this: https://gis.massdot.state.ma.us/arcgis/rest/services/Multimodal/GTFS_Systemwide/MapServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json

MBTAStops <- st_read(github_url) %>% 
  st_transform(st_crs(tracts16))  
```

```{r}

#clipping to tracts boundaries 

clippedMBTA <- 
  st_intersection(MBTAStops, tracts16)

#plotting

ggplot() + 
  geom_sf(data=st_union(tracts16)) +
  geom_sf(data=clippedMBTA) +
  labs(title="MBTA Stops", 
       subtitle="Boston, MA", 
       caption="Figure 1.1") +
  mapTheme()
```

```{r}
#creating a buffer
MBTABuffers <- 
  rbind(
    st_buffer(clippedMBTA, 2640) %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
    st_union(st_buffer(clippedMBTA, 2640)) %>% #the number at the end is in feet, because our CRS is in feet
      st_sf() %>%
      mutate(Legend = "Unioned Buffer")) #all st functions will be spatial related stuff

ggplot() +
  geom_sf(data=MBTABuffers) +
  geom_sf(data=clippedMBTA, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 1.2") +
  mapTheme()

#just the buffers

buffer <- filter(MBTABuffers, Legend=="Unioned Buffer")

```

```{r}

#selecting the TOD tracts 

selectCentroids <-
  st_centroid(tracts16)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts16, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")

ggplot() +
  geom_sf(data=selectCentroids, aes(fill = TotalPop)) +
  geom_sf(data=clippedMBTA, show.legend = "point") +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()

```

```{r}

#comparing TOD vs non TOD indicators 

allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2016", MedRent * 1.10, MedRent)) 



```

Can you re-create this figure using three geom_sf layers? The first is a basemap; the second maps rents using fill = q5(MedRent.inf), removing tract boundaries by setting colour=NA; and the third overlays buffer, setting colour = "red" and fill = NA.

```{r}
#plotting total pop

ggplot()+
  geom_sf(data = allTracts.group, aes(fill = q5(TotalPop)), color="transparent")+
   scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "TotalPop"),
                    name = "Popluation\n(Quintile Breaks)") + mapTheme() +
  facet_wrap(~year)+
  geom_sf(data = allTracts.group %>%
            filter(TOD == "TOD") %>%
            st_union(),
          color = "red",
          fill = "transparent")+
    
  labs(
    title = "Total Population, 2016-2020",
    subtitle = "",
    caption = "Figure Number, Data: US Census Bureau, ACS 5-year estimates",
    fill = "Total Population")
  
```

```{r}
#plotting med inc
ggplot()+
  geom_sf(data = allTracts.group, aes(fill = q5(MedInc)), color="transparent")+
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "MedInc"),
                    name = "Median Income\n(Quintile Breaks)") + mapTheme() +
  facet_wrap(~year)+
  geom_sf(data = allTracts.group %>%
            filter(TOD == "TOD") %>%
            st_union(),
          color = "red",
          fill = "transparent")+
    
  labs(
    title = "Median Income in 2020 adjusted Dollars, 2016-2020",
    subtitle = "",
    caption = "Figure Number, Data: US Census Bureau, ACS 5-year estimates",
    fill = "Median Income")
  
```

```{r}
#plotting med rent
ggplot()+
  geom_sf(data = allTracts.group, aes(fill = q5(MedRent)), color="transparent")+
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "MedRent"),
                    name = "Median Rent\n(Quintile Breaks)") + mapTheme() +
  facet_wrap(~year)+
  geom_sf(data = allTracts.group %>%
            filter(TOD == "TOD") %>%
            st_union(),
          color = "red",
          fill = "transparent")+
    
  labs(
    title = "Median Rent in 2020 adjusted Dollars, 2016-2020",
    subtitle = "",
    caption = "Figure Number, Data: US Census Bureau, ACS 5-year estimates",
    fill = "Median Rent")
  
```
```{r}
#plotting pct income spent on rent
ggplot()+
  geom_sf(data = allTracts.group, aes(fill = q5(IncSpentOnRentPCt)), color="transparent")+
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "IncSpentOnRentPCt"),
                    name = "PCT_of_Inc_On_Rent\n(Quintile Breaks)") + mapTheme() +
  facet_wrap(~year)+
  geom_sf(data = allTracts.group %>%
            filter(TOD == "TOD") %>%
            st_union(),
          color = "red",
          fill = "transparent")+
    
  labs(
    title = "Percentage of Income Spent on Rent, 2016-2020",
    subtitle = "",
    caption = "Figure Number, Data: US Census Bureau, ACS 5-year estimates",
    fill = "Median Rent")
  
```

```{r}
#plotting BusArtSciences

ggplot()+
  geom_sf(data = allTracts.group, aes(fill = q5(BusArtsScience)),color="transparent")+
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "TotalPop"),
                    name = "Number of Residents\n(Quintile Breaks)") + mapTheme() +
  facet_wrap(~year)+
  geom_sf(data = allTracts.group %>%
            filter(TOD == "TOD") %>%
            st_union(),
          color = "red",
          fill = "transparent")+
  labs(
    title = "People Employed in the Business, Arts or Sciences Sector, 2016-2020",
    subtitle = "",
    caption = "Figure Number, Data: US Census Bureau, ACS 5-year estimates",
    fill = "Population")
  
```

```{r}

# tables 

allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>% #best to go non spatil here because we're just making a table
  group_by(year, TOD) %>%
  summarize(Population = mean(TotalPop, na.rm = T),
            Rent = mean(MedRent, na.rm = T),
            Income = mean(MedInc, na.rm = T),
            Bus_Art_Sciences = mean(BusArtsScience, na.rm = T),
            Percent_Inc_on_Rent = mean(IncSpentOnRentPCt, na.rm = T))

allTracts.Summary %>%
  unite(year.TOD, year, TOD, sep = ": ", remove = T) %>%
  gather(Variable, Value, -year.TOD) %>%
  mutate(Value = round(Value, 2)) %>%
  spread(year.TOD, Value) %>%
  kable() %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.3")
```

```{r}
#grouped bar graphs

allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~Variable, scales = "free", ncol=5) +
    scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
    labs(title = "Indicator differences across time and space") +
    theme(strip.text = element_text(size = 7))+
    theme(legend.position="bottom")
```

